---
title: "Take Home Exercise-3"
description: |
  Reveal the economic of the city of Engagement, Ohio USA by using appropriate static and interactive statistical graphics methods for vast 2022 challenge.
author:
  - name: Shachi Anirudha Raodeo 
    url: https://github.com/ShachiR/ISSS608
    affiliation: School of Computing and Information Systems
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      eval = TRUE,
                      message = FALSE)
```

## The Task
This take-home exercise aims to reveal the economic of the city of Engagement, Ohio USA by using visualization techniques in R.

Economic considers the financial health of the city. Over time, are businesses growing or shrinking? How are people changing jobs? Are standards of living improving or declining over time?

Consider the financial status of Engagement’s businesses and residents, and use visual analytic techniques to address these questions.

Over the period covered by the dataset, which businesses appear to be more prosperous? Which appear to be struggling? 

Links to the dataset:

[`CheckinJournal.csv`](https://vast-challenge.github.io/2022/)
[`TravelJournal.csv`](https://vast-challenge.github.io/2022/)

## Step-by-step Data Visualisation

### Installing and launching R packages

Packages, namely `tidyverse` and `ggplot2` are required for this exercise. This code chunk installs the required packages and loads them onto RStudio environment.


```{r}
packages = c('tidyverse','ggplot2','ggdist', 'ggridges','patchwork', 'ggthemes',                      'hrbrthemes','ggrepel','ggforce',"HH","vcd",'scales','grid','gridExtra',
             'formattable','readr', 'ggiraph', 'plotly', 'DT', 'gganimate','readxl',                  'gifski', 'gapminder','treemap', 'treemapify','rPackedBar','ggstatsplot','ggside','broom','crosstalk')
for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}
```

## Data Preparation

### Data Source
The dataset used in this exercise is Participants.csv, published by the IEEE for [VAST challenge 2022] (https://vast-challenge.github.io/2022/) 

### Importing the dataset

The code chunk below imports *Restaurants.csv* and *TravelJournal.csv* and *Pubs.csv* from the data folder into R by using [`read_csv()`](https://readr.tidyverse.org/reference/read_delim.html) function of **readr** and saves it as Tibble data frame called *travel* and *restaurants* and *pubs*



```{r}
restaurants <- read_csv("data/Restaurants.csv")
summary(restaurants)
```
```{r}
pubs <- read_csv("data/Pubs.csv")
summary(pubs)
```
```{r eval=FALSE}
travel <- read_csv("data/TravelJournal.csv")
summary(restaurants)
```
### Missing value check

Before we proceed with data visualization we check if the datasets chosen  have any null values using the code below.

```{r}
apply(restaurants, 2, function(x) any(is.na(x)))

```
```{r}
apply(pubs, 2, function(x) any(is.na(x)))

```

```{r eval=FALSE}
apply(travel, 2, function(x) any(is.na(x)))

```

Since there are no null values observed we proceed with our next step.


## Data Wrangling 

The Travel Journal contains financial transactions by a participant towards Work/Home Commute, Eating, Coming Back From Restaurant,Recreation (Social Gathering), Going Back to Home. 
*We filter out the records related to Eating and Recreation (Social Gathering)*.

```{r eval=FALSE}
travel_filt <- travel[travel$purpose %in% c("Eating","Recreation (Social Gathering)"),]
```

### Calculating Amount Spent

Calculating *the total amount spent at the location as a difference of the starting balance and ending balance* in the travel journal

```{r eval=FALSE}
travel_filt$amountSpent <- travel_filt$startingBalance -travel_filt$endingBalance
```

```{r eval=FALSE}
saveRDS(travel_filt, 'data/travel_filt.rds')
```

```{r}
travel_filt <- readRDS('data/travel_filt.rds')
head(travel_filt)
```

## Data Visualization


Grouping the data by the travelEndLocationId which is equal to the restaurant ID or the pub ID.

```{r}
travel_group = travel_filt %>%group_by(travelEndLocationId) %>%
  summarise(amountSpent = sum(amountSpent), 
            .groups = 'drop')%>%
  arrange(desc(amountSpent))

``` 

```{r}

library(dplyr)
data_merge <-merge(x=travel_group, y=restaurants, by.x = 'travelEndLocationId', by.y =  'restaurantId')
data_merge$travelEndLocationId <- as.character(data_merge$travelEndLocationId)
data_merge$amountSpent <- data_merge$amountSpent/1000
```


above plot is modified considering the principles of data visualization. The above plot has simple X and Y axis, and is not so visually appealing. A line has been plotted along with the bar chart to make the differences in count easily visible to the reader.


```{r}
ggplot(data = data_merge,aes(x = travelEndLocationId, y=amountSpent)) +
 geom_col()+
  xlab("Business ID") +
  ylab("Revenue (in thousands $)") 
 
```

### Revenue for Restaurants

The restaurants highlighted in red are among those which had a revenue less than 50 thousand dollars over the period of time


```{r}
library(plotly)

color = c('rgba(222,45,38,0.8)','rgba(204,204,204,1)','rgba(204,204,204,1)', 'rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)', 'rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(222,45,38,0.8)','rgba(222,45,38,0.8)','rgba(222,45,38,0.8)',
          'rgba(222,45,38,0.8)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)')

fig<- plot_ly(data_merge, x = ~reorder(`travelEndLocationId`, -amountSpent), y = ~amountSpent, type = 'bar', marker = list(color = color))

fig <- fig %>% layout(title = "Revenue for Restraunts",
         xaxis = list(title = "Retaurant ID"),
         yaxis = list(title = "Revenue (in thousands $)"))

fig

```

### Revenue for Pubs

The restaurants highlighted in red are among those which had a revenue less than 300 thousand dollars over the period of time

```{r}
data_pub <-merge(x=travel_group, y=pubs, by.x = "travelEndLocationId", by.y = "pubId")
data_pub$travelEndLocationId <- as.character(data_pub$travelEndLocationId)
data_pub$amountSpent <- data_pub$amountSpent/1000

```

```{r}
library(plotly)
color = c('rgba(222,45,38,0.8)','rgba(222,45,38,0.8)','rgba(222,45,38,0.8)', 'rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)','rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)', 'rgba(204,204,204,1)')
fig<- plot_ly(data_pub, x = ~reorder(`travelEndLocationId`, -amountSpent), y = ~amountSpent, type = 'bar', marker = list(color = color))

fig <- fig %>% layout(title = "Revenue for Pubs",
         xaxis = list(title = "Pub ID"),
         yaxis = list(title = "Revenue (in thousands $)"))

fig
```
```{r}
treemap(data_merge,
        index=c("travelEndLocationId"), 
        vSize="amountSpent",
        vColor="amountSpent",
        title="Amount Spent in thousands of Dollars - Restaurants",
        title.legend = "Amount Spent in thousands of Dollars - Restaurants"
        )
```

```{r}
library(ggdist)
travel_filt %>%
  ggplot(aes(x = purpose, 
             y = amountSpent)) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
   theme_bw()+
  scale_y_continuous(breaks = seq(0, 35, by = 5), 
                     limits = c(0, 35))+
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")+
    labs(
    y= 'Revenue',
    x= 'Purpose',
    title = "Revenue generated by Purpose",
    caption = "Economic information, Ohio USA"
  )+
  theme(
    axis.title.y= element_text(angle=0), axis.ticks.x= element_blank(),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 1),
    
  )
```


 